k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-01-31 paper.Rmd
starting new paper

---
title             : "k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity"
shorttitle        : "k-plus anticlustering"

author:
  - name          : "Martin Papenberg"
    affiliation   : "1"
    corresponding : yes
    address       : "Heinrich-Heine-Universität Düsseldorf,  Institut für Experimentelle Psychologie, Universitätsstraße 1, 40225 Düsseldorf, Germany"
    email         : "martin.papenberg@hhu.de"

@@

  Martin Papenberg, Department of Experimental Psychology, Heinrich Heine University Düsseldorf.

@@

    Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---and is usually approached by maximizing instead of minimizing a clustering objective function. This paper presents k-plus, an extension of the classical k-means objective to maximize between-group similarity in anticlustering applications. K-plus represents between-group similarity as discrepancy in distribution moments (means, variance, and higher order moments), whereas the k-means criterion only reflects groups differences with regard to means. While constituting a new criterion for anticlustering, it is shown that k-plus anticlustering can be implemented by optimizing the original k-means criterion after the input data has been augmented with additional variables. A computer simulation and practical examples show that k-plus anticlustering achieves high between-group similarity with regard to multiple objectives. In particular, optimizing between-group similarity with regard to variances usually does not compromise similarity with regard to means; the k-plus extension is therefore generally preferred over classical k-means anticlustering. Examples are given on how k-plus anticlustering can be applied to real norming data using the open source R package anticlust, which is freely available via CRAN.

@@

keywords          : "Anticlustering, k-means, k-plus, variance, skewness, kurtosis, stimulus selection"
wordcount         : "6825"

@@

output            : 
  papaja::apa6_pdf
appendix: "appendix.Rmd"

@@

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
note              : "\\begin{flushleft} This paper is not the copy of record and may not exactly replicate the final, authoritative version of the article. The final article is available from the \\emph{British Journal of Mathematical and Statistical Psychology} via its DOI: \\href{https://doi.org/10.1111/bmsp.12315}{10.1111/bmsp.12315} \\end{flushleft}"
---

```{r setup, include = FALSE}
library(papaja) # use the development version: `remotes::install_github("crsh/papaja@devel")`; currently 0.1.1.9001
library(anticlust)
library(knitr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(DescTools)
# for formatting numbers in Rmarkdown, use my package prmisc (https://github.com/m-Py/prmisc)
library(prmisc) # remotes::install_github("m-Py/prmisc") 
options(digits = 2)
knitr::opts_chunk$set(message=FALSE, warning = FALSE, echo = FALSE, dev = "cairo_pdf")

@@

# Problem formalization

For the purpose of problem formalization, I adopt the notation @steinley2006 provided in his comprehensive synthesis of half a century of research on k-means. K-means anticlustering is used to partition $N$ elements each having $P$ attributes into $K$ groups $(C_1, \ldots, C_K)$. The partitioning is subject to the requirement that each element is assigned to exactly one group. In anticlustering applications, an additional constraint is usually imposed on the group sizes $n_k$ ($k = 1, \ldots K$), with the most common restriction being that all groups have equal size, i.e. $n_k = \frac{N}{K}$ [@papenberg2020].

For a data matrix $\mathbf{X}_{N \times P} = \{x_{ij}\}_{N \times P}$, k-means anticlustering aims to maximize the error sum of squares ($\mathit{SSE}$), which is the sum of the squared Euclidean distances between each data point and its cluster centroid:

$$
\mathit{SSE} = 
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{x}_j ^{(k)})^2
$$

The *k*'th centroid $\overline{\mathbf{x}}^{(k)} = (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})$ is a vector of length $P$, where each entry is the mean value of one of the $P$ attributes, computed across all observations belonging to the $k$'th cluster $C_k$ ($k = 1, \ldots, K$):

$$
\overline{x}_j ^{(k)} = \frac{1}{n_k} \sum\limits_{i \in C_k} x_{ij}
$$

Note that there is a direct connection between the $\mathit{SSE}$ and the location of the cluster centroids [@spath1986]. If the $\mathit{SSE}$ is minimal---which is the goal of k-means *clustering*---, the centroid values $\overline{\mathbf{x}}^{(k)} (k = 1, \ldots, K)$ are as far away as possible from the overall data centroid $\overline{\mathbf{x}} = (\overline{x}_1, \overline{x}_2, \ldots, \overline{x}_P)$, where $\overline{x}_j$ ($j = 1, \ldots, P$) is the overall mean value on the $j$'th attribute:

$$
\overline{x}_j = \frac{1}{N} \sum\limits_{i=1}^{N} x_{ij}
$$

If the $\mathit{SSE}$ is maximal---which is the goal of k-means *anticlustering*---, the cluster centroids are as close as possible to the overall centroid $\overline{\mathbf{x}}$ and therefore to each other [@spath1986]. Thus, k-means anticlustering directly optimizes the similarity of the mean attribute values between clusters.

## Minimizing discrepancy in variance between groups

In the following, I present a variation of the $\mathit{SSE}$ that can be used to minimize differences in variances between groups. To motivate the new criterion, we first consider a set of one-dimensional observations. The sample variance of a set of $N$ observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{x}$:

$$
\mathit{Var(x)} = \frac{1}{N} \sum\limits_{i=1}^{N} (x_i - \overline{x})^2
$$

When defining a new variable $y = (y_1, \ldots, y_N)$ as the squared deviation of each observation from the mean $\overline{x}$, i.e. $y_i = (x_i - \overline{x})^2$, the sample variance of $x$ can be reformulated as the mean of $y$:

$$
\mathit{Var(x)} = \frac{1}{N} \sum\limits_{i=1}^{N} y_i
$$

Because the sample variance is basically defined as an average---and k-means anticlustering minimizes differences with regard to averages---k-means anticlustering can be employed to equalize the variance across groups. This constitutes the central insight motivating the new k-plus criterion.

Thus, to obtain an adaptation of the $\mathit{SSE}$ whose maximization can minimize differences in variances between groups, a new variable is computed for each original variable: The new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ with $y_{ij} = (x_{ij} - \overline{x}_j)^2$. On the basis of $\mathbf{Y}_{N \times P}$, we define the criterion $\mathit{SSE_{Var}}$:

$$
\mathit{SSE_{Var}} = 
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{y}_j ^{(k)})^2
$$

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ can be used to obtain groups that have similar variances across the $P$ attributes. As a caveat, however, it is crucial to note that maximizing $\mathit{SSE_{Var}}$ only reliably yields similar group variances if the group centroids $\overline{\mathbf{x}}^{(k)}$ ($k = 1, \dots, K$) are close to the overall data centroid $\overline{\mathbf{x}}$. This is because the within-group variance is calculated on the basis of the group centroid $\overline{\mathbf{x}}^{(k)}$, but the new variables $y_{ij}$ are calculated using the overall centroid $\overline{\mathbf{x}}$ as reference point. Using $\overline{\mathbf{x}}$ as reference point is however necessary because the group means $\overline{\mathbf{x}}^{(k)}$ are determined by the composition of the groups $(C_1, \ldots, C_K)$, which is the end result of the optimization and not known in advance. To ensure that $\mathit{SSE_{Var}}$ is a valid criterion, we must therefore ensure that the group centroids of the original variables are close to the overall data centroid (and therefore to each other). Ensuring that the group means are close to the overall mean is accomplished by maximizing the classical $\mathit{SSE}$, i.e., via standard k-means anticlustering (Späth, 1986). Combining the maximization of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ at the same time leads to the k-plus approach for anticlustering, which is detailed in the next section.

## A bicriterion model: K-plus anticlustering {#kplus}

An exclusive maximization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is not feasible---and most likely not interesting on its own---without also considering the standard $\mathit{SSE}$ (i.e., equalizing means). The most simple approach for an optimization that considers multiple criteria at the same time is the weighted sum approach [@naidu2014; @marler2010]. That is, all criteria are computed independently from each other and are then combined into a single criterion through a weighted sum. By combining $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ in a single criterion, we define the objective function for k-plus anticlustering:

$$
\mathit{SSE_{kplus}} = 
w_1 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{x}_j ^{(k)})^2 + 
w_2 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{y}_j ^{(k)})^2
$$

```{r motivatingExample}

# some functions for motivating example

# Variance objective function with changed order of the arguments 
# (can be called by sapply() etc.)
var_objective <- function(clusters, features) {
  anticlust::variance_objective(features, clusters)
}

# Compute features for k-plus criterion
squared_from_mean <- function(data) {
  apply(data, 2, function(x) (x - mean(x))^2)
}

# First rendering is slow because the data needs to be generated
path <- "partition_objectives.csv"

if (!file.exists(path)) {
  set.seed(2207) # 2207
  # Generate data and all partitions
  N <- 14
  K <- 2
  partitions <- generate_partitions(N, K)
  features <- rnorm(N, 15, 3)

  # Compute objectives for each partition
  # For each partition: Compute SSE objective, and SSE(VAR)
  # 1. SSE
  all_objectives_kMeans <- sapply(
    partitions,
    FUN = var_objective,
    features = features
  )

  squared_distances_from_mean <- squared_from_mean(as.matrix(features))

  # 2. SSE(VAR)
  all_objectives_kVar <- sapply(
    partitions,
    FUN = var_objective,
    features = squared_distances_from_mean
  )

  # 3. SSE(KPLUS)
  all_objectives_kPlus <- sapply(
    partitions,
    FUN = var_objective,
    features = cbind(features, squared_distances_from_mean)
  )

  df <- data.frame(
    kmeans = all_objectives_kMeans,
    kvar = all_objectives_kVar,
    kplus = all_objectives_kPlus
  )

  write.table(
    df, path, row.names = FALSE, sep = ","
  )

  # investigate the best 0.5% of k-means anticlustering partitions, for discussion
  # best_0.1percent_anticlusterings <- df$kmeans /max(df$kmeans) > .999
  # best_partitions <- partitions[best_0.1percent_anticlusterings]
  # 
  # neighbor_matrix <- anticlust:::selection_matrix_from_clusters
  # neighbourhoods <- lapply(best_partitions, neighbor_matrix)
  # neighbourhoods <- lapply(neighbourhoods, function(x) as.integer(x[lower.tri(x)]))
  # pairs <- Reduce("+", neighbourhoods)
  # length(pairs)
  # length(neighbourhoods[[1]])
  # choose(14, 2)
  # min(pairs) # 40
  # max(pairs) # 95
  # length(best_partitions) # 146
  # -> all items are grouped with one another, and no items are always grouped with one another

} else {
  df <- read.csv(path)
}

By maximizing $\mathit{SSE_{kplus}}$, we seek to obtain a partitioning where means as well as variances are similar between groups, while the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of means and variances. However, research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict rather than in alignment [@ferligoj1992direct]. As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown if the two constituting criteria are aligned. That is, it has yet to be demonstrated that maximizing $\mathit{SSE_{kplus}}$ can yield a partition where means as well as variances are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatter plot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The criteria seem to be rather well aligned: There is a high concentration of partitions where both the $\mathit{SSE}$ and the $\mathit{SSE_{Var}}$ have high values. At the same time, there are many partitions that fulfil only one of the criteria, but fail to simultaneously fulfil the other criterion. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

```{r, fig.cap = "The relationship between the objectives $\\mathit{SSE}$ and $\\mathit{SSE_{Var}}$ across all ways to partition $N = 14$ normally distributed random values into $K = 2$ groups. The partition represented by the square maximizes $\\mathit{SSE}$ and thus minimizes the difference in means between groups; the partition represented by the triangular maximizes $\\mathit{SSE_{Var}}$. The partition represented by the diamond maximizes $\\mathit{SSE_{kplus}}$ using $w_1 = w_2 = 1$.", fig.height = 5, fig.width = 5}

# Plot SSE vs SSE(VAR)
plot(
  df[, c("kmeans", "kvar")], 
  pch = 4, col = "darkgrey", 
  las = 1, cex = .5,
  xlab = "",
  ylab = ""
)
mtext(side = 2, text = expression(italic(SSE[Var])), line = 3)
mtext(side = 1, text = expression(italic(SSE)), line = 3)

# Illustrate best partition wrt difference in means (i.e., SSE)
points(
  df$kmeans[which.max(df$kmeans)],
  df$kvar[which.max(df$kmeans)],
  cex = 1.2, pch = 22, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt difference in variance (i.e., SSE(VAR))
points(
  df$kmeans[which.max(df$kvar)],
  df$kvar[which.max(df$kvar)],
  cex = 1.2, pch = 24, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt k-plus criterion
points(
  df$kmeans[which.max(df$kplus)],
  df$kvar[which.max(df$kplus)],
  cex = 1.2, pch = 23, bg = "white", lwd = 1.7
)
```

## Generalizing k-plus anticlustering to higher order moments

The k-plus logic to equalizing group variances as well as group means can be seamlessly generalized to higher order moments. The *t*'th sample moment of a variable $x = (x_1, \ldots, x_N)$ can be computed as

$$
\frac{1}{N} \sum\limits_{i=1}^{N} (x_i - \overline{x})^t.
$$

where $t = 2$ gives the variance, $t = 3$ the skewness, and $t = 4$ the kurtosis [@joanes1998]. Thus, by changing $t$---the power of the deviation between mean and data points---we can use the k-plus logic to equalize any desired sample moment between groups. That is, to extend the k-plus criterion to higher order moments, we compute a new variable for each original variable and for each distribution moment that should be considered in the anticlustering process. Analogously to the computation of $\mathit{SSE_{Var}}$, we could then define an $\mathit{SSE_{Skew}}$ using $t = 3$ or an $\mathit{SSE_{Kurtosis}}$ using $t = 4$. However, to formalize the computation of the k-plus criterion for an arbitrary number of distribution moments, some new notation has to be provided.

We define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{\cdot j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean---the overall mean of the $j$'th attribute---is $\overline{x}_j$. Let $T$ be the number of distribution moments that should be similar across $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{\cdot j}$ ($j = 1, \ldots, P$), a new *k-plus variable* $\hat{x}_{t \cdot j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

$$
\hat{x}_{tij} = 
\begin{cases}
    x_{ij} & \text{if } t = 1\\
    (x_{ij} - \overline{x}_j)^t, & \text{if } t > 1.
\end{cases}
$$

This computation results in a three dimensional tensor $\mathbf{\hat{X}}_{T \times N \times P}$ = $\{\hat{x}_{tij}\}_{T \times N \times P}$ where---in comparison to the original data matrix $\mathbf{X}_{N \times P}$---the additional dimension is given by the additional distribution moments that are considered in the anticlustering process. When defining $\hat{x}_{tij}$, the condition $t = 1$ is separately included to keep the unchanged original data as first entry of the "moment dimension" in $\mathbf{\hat{X}}_{T \times N \times P}$. It is used to equalize the mean---the first moment---between groups. Using $\mathbf{\hat{X}}_{T \times N \times P}$ and a vector of weights $\mathbf{w} = (w_1, \dots, w_T$), the generalized multicriterion for k-plus anticlustering $\mathit{SSE_{kplusGeneralized}}$ can be formalized:

$$
\mathit{SSE_{kplusGeneralized}} =
  \sum\limits_{t=1}^{T} 
  w_t \left(
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (\hat{x}_{tij} - \overline{x}_{tj} ^{(k)})^2
  \right)
$$

Here, $w_t$ is the weight associated with the $t$'th moment and $\overline{\mathbf{X}}^{(k)}_{T \times P} = \{\overline{x}_{tj}^{(k)}\}_{T \times P}$ is a matrix containing the mean values for all k-plus variables in the $k$'th group, i.e.

$$
\overline{x}_{tj}^{(k)} = \frac{1}{n_k} \sum_{i \in C_k} \hat{x}_{tij}
$$

Maximizing $\mathit{SSE_{kplusGeneralized}}$ can be used to minimize differences with regard to the first $T$ distribution moments between $K$ groups.

## The unweighted sum approach for k-plus anticlustering

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$ by setting $w_t = 1$ ($t = 1, \dots, T$). This simplification leads to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$ can be dropped. Instead, the tensor is flattened into a two-dimensional matrix by appending the newly computed k-plus variables $\hat{x}_{tij}$ ($t \ge 2$) to the original data matrix $\mathbf{X}_{N \times P}$. We obtain the augmented matrix $\mathbf{X'}_{N \times \mathit{TP}}$:

$$
  \mathbf{X'}_{N \times \mathit{TP}} =
  \left[ {\begin{array}{ccc|ccc|c|ccc}
    x_{11} & \cdots & x_{1P} & 
        (x_{11} - \overline{x}_1)^2 & \cdots & (x_{1P} - \overline{x}_P)^2 & 
        \cdots & 
        (x_{11} - \overline{x}_1)^T & \cdots & (x_{1P} - \overline{x}_P)^T \\
    x_{21} & \cdots & x_{2P} & 
        (x_{21} - \overline{x}_1)^2 & \cdots & (x_{2P} - \overline{x}_P)^2 & 
        \cdots & 
        (x_{21} - \overline{x}_1)^T & \cdots & (x_{2P} - \overline{x}_P)^T \\
    \vdots & \ddots & \vdots & \vdots & \ddots & \vdots & \cdots & \vdots & \ddots & \vdots\\
    x_{N1} & \cdots & x_{NP} & 
        (x_{N1} - \overline{x}_1)^2 & \cdots & (x_{NP} - \overline{x}_P)^2 & 
        \cdots & 
        (x_{N1} - \overline{x}_1)^T & \cdots & (x_{NP} - \overline{x}_P)^T \\
  \end{array} } \right]
$$

When defining the rows of $\mathbf{X'}_{N \times \mathit{TP}}$ with the notation $x'_{i \cdot} = (x'_{i1}, \ldots, x'_{i(TP)})$ ($i = 1, \ldots, N$), the criterion $\mathit{SSE_{kplusGeneralized}}$ can then be computed as

$$
\mathit{SSE_{kplusGeneralized}} =
  \sum\limits_{j=1}^{\mathit{TP}}
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x'_{ij} - \overline{x'}_j ^{(k)})^2.
$$

Thus, when setting the weights $w_t = 1$ ($t = 1, \dots, T$), the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ can be expressed in the same form as the standard k-means criterion $\mathit{SSE}$. According to this principle, the k-plus criterion is computed by first creating the additional k-plus variables, appending them as new columns to the original data matrix, and then computing the standard $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $\mathbf{w}$ are ignored when computing $\mathit{SSE_{kplusGeneralized}}$. Because the computation of the new k-plus variables involves exponentiation, their range of values is expected to be very different from the original variables. This imbalance can strongly and unpredictably influence the relative weight of the different moments when computing $\mathit{SSE_{kplusGeneralized}}$ (e.g., see Figure 1). To ensure a comparable weight of all criteria, all attributes---i.e., the original data as well as the new k-plus variables---can be standardized before computing $\mathit{SSE_{kplusGeneralized}}$, e.g. via $z$-standardization. This approach will henceforth be referred to as the unweighted standardization approach for k-plus anticlustering.

@@

## Simulation 1: Selecting weights for k-plus anticlustering

```{r getresultsSim1}
dir <- "./Simulation_1/"
files <- list.files(dir, pattern = ".Rdata")
results <- list()
for (file in files) {
  results <- c(results, get(load(paste0(dir, file))))
}

Six weighting schemes were used to optimize $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, thus minimizing differences with regard to differences with regard to group means, variances, skewness and kurtosis. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The first unweighted approach implemented a $z$-standardization of all columns---the original as well as the augmented k-plus variables---of the input data before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. The second unweighted approach used the unstandardised augmented data matrix. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favouring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favouring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

In the simulation, `r length(results)` data sets from were generated from a normal distribution ($M = 0$, $\mathit{SD} = 1$). Each data set was divided into groups via optimization of $\mathit{SSE_{kplusGeneralized}}$, once using each of the six weighting schemes. The number of variables $P$ was determined randomly for each data set and varied between 1 and 5; the number of groups $K$ was also determined randomly for each data set and varied between 2 and 4; the sample size $N$ randomly varied between 20 and 100 with the restriction that $N$ had to be divisible by $K$.

```{r}
final_tab_sim1 <- round(Reduce("+", results) / length(results) * 100, 2)
rownames(final_tab_sim1) <- c(
  "$\\mathbf{w} = (1, 1, 1, 1)$ / unstandardized",
  "$\\mathbf{w} = (1, 1, 1, 1)$ / standardized",
  "$\\mathbf{w} = (10, 1, 1, 1)$",
  "$\\mathbf{w} = (1, 10, 1, 1)$",
  "$\\mathbf{w} = (1, 1, 10, 1)$",
  "$\\mathbf{w} = (1, 1, 1, 10)$"
)
colnames(final_tab_sim1) <- c(
  "$\\mathit{SSE}$",
  "$\\mathit{SSE_{Var}}$", 
  "$\\mathit{SSE_{Skew}}$",
  "$\\mathit{SSE_{Kurtosis}}$",
  "Average"
)
note <- paste0("Cells indicate how well the maximization of $\\mathit{SSE_{kplusGeneralized}}$ fulfilled each of the four constituting criteria $\\mathit{SSE}$, $\\mathit{SSE_{Var}}$, $\\mathit{SSE_{Skew}}$, and $\\mathit{SSE_{Kurtosis}}$, depending on the weighting scheme. For each simulation run, the relative performance was computed as the result that was obtained by the given weighting scheme, divided by the best result that was obtained by any weighting scheme in this simulation run. Cell values represent the average relative performance across all ", length(results), " simulation runs.")

apa_table(
  final_tab_sim1, 
  escape = FALSE, 
  caption = "Average relative performance of the competing weighting schemes in Simulation 1.",
  note = note,
  stub_indents = list("Unweighted approaches" = 1:2,
                      "Weighted approaches" = 3:6)
)

unstd_best <- sapply(results, function(x) names(which.max(x[, 5]))) == "unweighted_std"

# significance test
sig_test <- prop.test(x = sum(unstd_best), n = length(unstd_best), p = 1/6)

Table 1 illustrates how well each of the six weighting schemes fulfilled each of the four constituting criteria, averaged across the `r length(results)` simulation runs. For the weighted approaches, the criterion that had received the highest weight during the optimization was fulfilled to the highest degree, as should be expected. The unweighted standardization approach yielded the best average performance across the four criteria. While the degree of advantage by itself does not seem to be remarkable, this pattern was observed in an astonishing `r mean(unstd_best) * 100`% of all simulation runs, while the level of chance would have been `r 1/6 * 100`% (`r format_p(sig_test$p.value)`). Among the competing weighting schemes, the unweighted standardization approach therefore yielded the most promising results. It performed also remarkably strong with regard to the standard $\mathit{SSE}$, even slightly outperforming the weighting scheme that explicitly favoured the $\mathit{SSE}$. Hence, the unweighted standardization approach implicitly gave high priority to minimizing differences in means. This is a pleasant result because, as discussed earlier, for the higher order $\mathit{SSE}$s to be valid, the group means have to be close to the overall mean, which is best accomplished if the standard $\mathit{SSE}$ has high priority.

## Simulation 2: Comparing k-plus to other anticlustering objectives

In Simulation 2, the performance of two k-plus criteria was compared to the performance of two traditional anticlustering criteria. "Standard" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplus}}$, thus minimizing differences with regard to group means and variances. "Generalized" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplusGeneralized}}$ using $T = 4$, thus minimizing differences with regard to group means, variances, skewness and kurtosis. The unweighted standardization approach that performed best in Simulation 1 was employed. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, and also (iii) a random allocation of elements to groups.

### Conditions

@@

### Evaluation criteria

After the four anticlustering methods and the random allocation procedure had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I investigated how well the competing methods were able to create similar groups. To this end, I analysed the discrepancy in means, standard deviations, skewness and kurtosis between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$).

As a secondary purpose of Simulation 2, I investigated the similarity of partitions obtained when optimizing the different anticlustering objectives. In cluster analysis, partitions identified through different approaches tend to overlap considerably, because similar elements tend to be placed in the same cluster while dissimilar elements tend to be placed in different clusters; this principle holds regardless of the specific clustering method. However, it is an open question whether the optimization of different anticlustering objectives leads to similar groupings of the elements in anticlusters. To address this question, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each simulation run, i.e., for data set and for each $K$.

### Results

Table 2 displays the global simulation results aggregated across all conditions. Standard k-means performed best at minimizing discrepancy with regard to means ($\Delta \mathit{M}$), but did not outperform a random assignment with regard to all other distribution characteristics. K-plus anticlustering addressed $\Delta \mathit{M}$ almost as well as k-means, and at the same time minimized discrepancy with regard to the standard deviations ($\Delta \mathit{SD}$); both k-plus approaches strongly outperformed all other methods with regard to minimizing $\Delta \mathit{SD}$.

\begin{singlespace}

```{r Table1}

# Load required packages
library(dplyr)
library(ggplot2)
library(tidyr)

## Analyze data for K = 2 and K = 3 and K = 4
simulation_results <- list()
for (K in 2:4) {
  filename <- paste0("./Simulation_2/results-K", K, "-objectives-raw.csv")
  df <- read.csv(filename, sep = ";", stringsAsFactors = FALSE)
  df$K <- K
  simulation_results[[paste0("K-", K)]] <- df
}

df <- do.call(rbind, simulation_results)
rownames(df) <- NULL

# Make long format
ldf <- pivot_longer(
  df,
  cols = paste0(c("means", "sd", "skew", "kur"), "_obj"),
  names_to = "Objective",
  names_pattern = "(.*)_obj"
)

ldf <- ldf[ldf$method != "k-plus-correlation", ] # removed because this deserves closer attention than what was given in the paper; check out the first version of the preprint for the previous results

ldf$method[ldf$method == "k-plus-skew-kurtosis"] <- "k-plus generalized"

## Global results, aggregated across all simulation variables
tab <- ldf %>% 
  group_by(method, Objective) %>% 
  summarise(Mean = round(mean(value), 3)) %>% 
  pivot_wider(names_from = Objective, values_from = Mean) %>% 
  select(c(means, sd, skew, kur))

colnames(tab) <- c("",
  "$\\Delta M$", 
  "$\\Delta \\mathit{SD}$",
  "$\\Delta \\mathit{Skew}$",
  "$\\Delta \\mathit{Kurtosis}$"
)

>>>>

---
title             : "An improved anticlustering method to maximize stimulus dispersion in experiments"
shorttitle        : "Maximum stimulus dispersion"

author:
  - name          : "Martin Papenberg"
    affiliation   : "1"
    corresponding : yes
    address       : "Heinrich-Heine-Universität Düsseldorf, Institut für Experimentelle Psychologie, Universitätsstraße 1, 40225 Düsseldorf, Germany"
    email         : "martin.papenberg@hhu.de"
  - name          : "Martin Breuer"
    affiliation   : "1"
    corresponding : no
  - name          : "Max Diekhoff"
    affiliation   : "1"
    corresponding : no
  - name          : "Gunnar W. Klau"
    affiliation   : "1"
    corresponding : no

@@

  Martin Papenberg, Department of Experimental Psychology, Heinrich 
  Heine University Düsseldorf. Martin Breuer, Gunnar W. Klau, Max Diekhoff, 
  Department of Computer Science, Heinrich Heine University Düsseldorf.

@@

    When assigning stimuli to sets in experimental psychology, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually substandard when done manually by researchers, automated algorithmic approaches are preferred to accomplish this goal. Brusco et al. (2020) proposed a bicriterion anticlustering algorithm (BILS) that simultaneously maximizes the diversity (corresponding to between-set similarity) and the dispersion (corresponding to pairwise stimulus dissimilarity within-sets). We provide a first independent assessment of BILS and present BILS-E, which extends BILS in three ways. First, a globally optimal value of dispersion is always returned. Second, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices, for example to include cannot-link constraints in the anticlustering process. Third, BILS-E allows that other anticlustering objectives of between-group similarity, such as k-means or k-plus, can be optimized instead of only the diversity. Using our optimal approach as benchmark in a simulation study, we show that the original BILS identifies optimal dispersion values for much larger data sets than previously analyzed, but also highlight conditions where it fails to do so. A practical example shows how BILS-E can be used for stimulus assignment in experimental psychology. The original BILS as well as our extension are both freely available via the open source R package anticlust.

@@

keywords          : "Anticlustering, maximum dispersion, optimal algorithm, bicriterion optimization"
wordcount         : "12345"

@@

output            : 
  papaja::apa6_pdf

@@

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"

---

```{r setup, include = FALSE}
library(papaja) # use the development version: `remotes::install_github("crsh/papaja@devel")`; currently 0.1.1.9001
library(anticlust)
library(knitr)
# library(dplyr)
# library(ggplot2)
# library(tidyr)
# library(DescTools)
# for formatting numbers in Rmarkdown, use my package prmisc (https://github.com/m-Py/prmisc)
library(prmisc) # remotes::install_github("m-Py/prmisc") 
options(digits = 2)
knitr::opts_chunk$set(message=FALSE, warning = FALSE, echo = FALSE, dev = "cairo_pdf")

@@

@@

## Conditions

@@

## Evaluation criteria

## Results
